Extended Early Adulthood and Attitude Change

### 2024-06-28 manuscript/formative_period_kk.Rmd
drafting, some changes to demo cleaning

Adolescence and young adulthood are frequently identified in social science literature as "impressionable years" -- a stage of the life course in which people are more susceptible to broad social currents, when their opinions are more malleable, and when they form the core ideas and attitudes that shape subsequent decision making [@krosnick1989; @alwin1991]. In support of this idea, social science research regularly finds that people in these life stages are more likely to change their political preferences, social attitudes, and cultural tastes than people in other life stages [@sears2013; @kiley2020]. This pattern, in which young people show greater openness to cultural change but become more resistant over time, underpins dynamics around cohortization that are frequently invoked in larger models of social and cultural change [@ryder1965; @mannheim1952].

Why people in this life stage appear to show more frequent opinion change relative to older people, however, is unclear. Explanations in the social sciences tend to highlight the unique social features of these early life stages such as high rates of mobility between social contexts, participation in educational institutions, and cultural expectations about pursuing "new experiences,” arguing that these conditions shape how people are exposed to new sources of information and their ability to update opinions in the face of new information [@alwin1991; @danigelis2007]. Other related explanations focus on the unique features of middle age that might stabilize attitudes during this period, including people holding positions with higher levels of social authority during middle age [@eaton2009; @visser2004]. At the same time, work in biology, psychology, and cultural evolution finds that peoples' cognition changes in ways that might make them more resistant to new information as they age [CITATION], suggesting the differences between early adults and the rest of the population might be more rooted in biological and cognitive mechanisms than social mechanisms.

While both social and cognitive/biological mechanisms likely play a role in facilitating opinion change during early adulthood, the relative contributions of each are unclear, and clarifying their relative importance is important for understanding the mutually reinforcing processes of social change and cultural evolution. If openness to cultural change in early adulthood is principally a function of the social structure of life during these ages, then delays in the timing and ubiquity of life course transitions like marriage and childbearing, and increases in college-going, could have significant effects on the broader process of macro-social change. Similarly, social changes such as the deinstitutionalization of marriage [@cherlin2004; @cherlin2021; @robbins2022] that decrease distinctions between life course stages might make older people more open to change in more recent eras than they were in earlier eras, diminishing the importance of "cohorts" for social change and increasing the importance of intrapersonal change over time [@kiley2020; @vaisey2016]. If the mechanisms are principally cognitive and biological, rather than social, then these broader changes are unlikely to affect rates of social change in the same way.

In this paper, we explore how rates of opinion change in different age groups have changed over the past 70 years in response to broader social structure and changes in the timing of life course events. Since the 1950s, the timing of life-course transitions such as completing schooling, entering the workforce, getting married, leaving one's parents' home, and having children have changed dramatically, with most of these event occurring later in the life course for more recent cohorts [@buchmann1989; @bruckner2005]. Similarly, the "early adulthood" stage of life has become more detached from family and other role obligations, giving people in this life stage more freedom to explore new ideas and lifestyle choices [@rosenfeld2009]. These broad social changes potentially lengthen of and intensify this stage of the life course in ways that potentially have consequences for the rate at which people undergo cultural changes.

Drawing on data for 77 questions asked multiple times across 12 panel surveys dating from the 1950s through the 2020s, we model the amount of change people make over time in social and political attitudes as a function of the year they were observed and their age at the time. Consistent with the "impressionable years" hypothesis, we find that members of the youngest age groups (18-25 and 26-33) have more variable opinions than older respondents. However, these differences are not constant over time. They emerged over the second half of 20th Century as middle-aged and older respondents came to report more stable opinions over time, while early adult respondents’ reported opinions continued to change at roughly the same rate over time.

Our findings complicate the narrative about social change and cultural change and raise questions about which social forces emerged to stabilize attitudes in middle age for more recent cohorts.

@@

These aggregate models of social imprinting have implications for the kinds of patterns we should observe in within-person opinion change over time. The "impressionable years" hypothesis -- that people demonstrate more opinion change early in life followed by a period of relative stability, especially in middle age -- has often been tested against other theories of opinion change over the life course. The most common alternatives are an "aging stability" or "increasing persistence" hypothesis that argues that openness to change gradually declines over time and a "life stages" hypothesis, in which people are open to attitude change both early in life and later in life, with stability principally happening in middle ages [@visser2004a]. While different attitudes occasionally follow different patterns, the impressionable years hypothesis has generally been shown to be true for broad political attitudes including political ideology, partisan identification, interest in politics, and specific policy preferences [@krosnick1989; @alwin1991; @sears1999; @prior2010]; social attitudes around race, gender, and sexuality [@danigelis2007]; as well as things like personality traits [@wagner2019] and cultural tastes [@]. More generally, recent research has found that, with a few notable exceptions of high-salience events such as national shifts around gay rights, durable change of social opinion is quite rare among U.S. adults [@kiley2020], and Vaisey and Lizardo [-@vaisey2016] find that cohort, the year of a person's birth, explains much more variance than period, the year of observation, for most items in the General Social Survey, suggesting a strong impressionable years pattern.

@@

While we have focused on changes in the timing and structure of life course stages over the past 70 years as our primary focus, the past 70 years have also seen significant shifts in the broader social structure and cultural environment that might affect attitude stability independent of changes in life course structure. We briefly consider three here.

First, the number of people pursuing and completing higher education increased significantly over the window studied here []. Existing work suggests that people with higher levels of education are more likely to express the same opinion over time in survey contexts [@alwin1991a; @alwin2007]. A population with higher levels of education might be more likely to hold stable opinions regardless of broader social transformations.

Second, technological changes have dramatically lowered barriers to accessing information, including information about social and political affairs that might facilitate stable opinion holding. Both broadcast and cable television and then through the internet and social media. People have greater access to information with which to form opinions.

Third, the past 70 years have seen extensive elite polarization around political and social affairs, as well broader public partisan alignment [@fiorina2008; @baldassarri2008]. People now receive clearer signals from cultural elites, including politicians, the news media, and others, about which beliefs and identities "go together" in the population, resulting in what DellaPosta (2020) calls an "oil spill" of polarization [@dellaposta2020]. This issue and identity alignment might make it easier for the public to identify what beliefs people like them "should" hold and hold them consistently. Geographic and associational sorting into lifestyle “enclaves” also reinforce these processes, continually providing clearer signals about which opinions people should hold [].

Broadly speaking, these three forces suggest that opinions . Additionally, peoples’ opinions on issues might

@@

Our general question is whether, when presented with the same question during the same age range, more recent cohorts will demonstrate more change over time than members of earlier cohorts. Analyzing this question requires data on opinion and attitude reports over time for members of different cohorts at different time points, with respondents coming from the same general population. To construct such a data source, we combine several panel surveys conducted by the American National Election Studies and the General Social Survey spanning about 60 years. Both the ANES and GSS are designed to generalize to the non-institutionalized U.S. adult (18+) population, and therefore can be thought of as tapping the same general population. We included a total of 12 panels in our search: ANES panels from 1956-60, 1972-76, 1980, 1990-92, 1992-97, 2000-04, 2016-20, and 2020-22; and GSS panels from 2006-10, 2008-12, 2010-14, and 2016-20. We evaluated an additional panel conducted by the ANES in 2008-10, but no questions overlapped this panel and any other panel.

For reasons explained below, we identified questions that were asked in at least three different panels with at least one of those panels asking some respondents to answer the question at three or more separate time points. Because three of the GSS panels overlap considerably (2006-10, 2008-12, and 2010-14) and contain the same set of questions, we did not treat these as separate panels for the purposes of generating three distinct time points. In other words, questions needed to appear in these panels and an ANES panel to be included in our analysis. When question wording changed across panels, we harmonize response options to the question structure with the fewest response options.

Our search produced a total of 77 questions tapping a broad range of topics including political and ideological identification, general government policy preferences, preferences for federal spending on different priorities, views on government efficacy, views on social issues and social change, and sentiment toward various political and social groups.

@@

@@

When questions are observed in multiple panels, differences in the amount of change we observe could be attributable to either changes in the population's capacity to report stable attitudes over time (our main interest) or to the structural features of the survey administration. Specifically, the amount of change observed over time might be due to variation in the duration between waves, with longer duration between observations leading to more wave-to-wave change as people experience more shocks that should shift their opinions. In our data, panel duration varied from less than one year (the 1980 ANES Major Panel Study) to about five years (the 1992-97 ANES), and duration between questions ranged from a few days (for questions asked in both pre- and post-election surveys in the ANES panels) to about five years (questions asked in 1992 and 1997), with peaks around 2 years and 4 years.

@@

In this equation, $d_{i}$ captures the duration of time between $t_1$ and $t_2$ in years for each pair of observations. Importantly, respondents might have multiple observation pairs for each panel, depending on the number of times a question is asked. In the 1972-76 ANES panel, respondents were asked their partisan identification on three occasions: in the 1972 pre-election survey, in the 1974 pre-election survey, and the 1976 pre-election survey, each about two years apart. If a person participates in all three waves, they have three observations of $|y_{2i}-y_{1i} |$ and $d_i$: 1972 and 1974 (about 2 years), 1974 and 1976 (about 2 years), and 1972 and 1976 (about 4 years).

Variation in duration between observations within panels lets us separate two kinds of change: 1) change attributable to durable shifts of opinions over time that can be associated with duration between observations and 2) change that is principally random, or unrelated to the duration between observations. If respondents have a stable underlying partisan identification they report with random error, and no real change occurs, then in expectation, duration should be unrelated to absolute change in responses over time, and $\beta_1 = 0$. In this formulation, $\beta_0$ captures the expected absolute change in responses over time when the duration between waves is 0, or, more intuitively, the random deviation between waves that is unrelated to durable change (random noise). Conversely, $\beta_1$ captures the expected change in absolute difference between waves for a one-unit (year) change in duration of observation, or the amount of absolute change we would expect over-and-above the change driven by random noise ($\beta_1$). For example, if a respondent shifts from "strong Democrat" (point 0 on the scale) to "Independent" (point 3 on the scale) between 1972 and 1974, and stays at "Independent" for the 1976 wave, they will have differences of 3 and 0 for the two-year duration (an average of 1.5) and 3 for the four-year duration, generating a positive coefficient for $\beta_1$ if this kind of switching behavior is common in the populations.

We can extend equation to incorporate age groups:

$$ |y_{2i}-y_{1i} | = \beta_0 + \beta_1*d_i + \beta_2(age_{26-33}) + \beta_3(age_{34-64}) + \beta_4(age_{65+}) + \beta_5(d*age_{26-33}) + \beta_6(d*age_{34-64}) + \beta_7(d*age_{65+}) + \epsilon_i $$

@@

$\beta_0$ through $\beta_7$ retain their general meaning as outlined above, but apply to time = 0, which we set to be the most recent observation date (2016). $\beta_8$ estimates the change in the baseline rate of change for the reference group. $\beta_9$ captures change in the effect of duration of observation over time for people in the reference group (ages 18-25). Coefficients $\beta_{10}$ through $\beta_{12}$ and $beta_{13}$ capture how differences in the baseline rate of change and duration effects between the reference group and older age categories change over time.

@@

$$ \beta_{0j} = \gamma_0 + \upsilon_{0j} $$
$$ \beta_{1j} = \gamma_1 + \upsilon_{1j} $$
$$ \beta_{2j} = \gamma_2 + \upsilon_{2j} $$
$$ \beta_{3j} = \gamma_3 + \upsilon_{3j} $$

This model allows for each question its own set of coefficient estimates weighted by the amount of information we have about the particular question. Questions asked in many panels

In practice, the model presented in Eq ### is likely too complicated in the face of the data limitations we face.

In this framework, the first-level equation models the absolute difference in a pair of responses to the same question given by the same respondent, similar to what was presented above. These equations are nested at the level of the survey item. Each item is assumed to have its own intercept (expected variance at time = 0 and duration = 0) and effects for duration ($beta_{1j}$), time ($\beta_{2j}$), and their interaction ($\beta_{3j}$). These effects are assumed to be normally distributed around a main effect. In practice, we

# Weights

@@

## Demographic Variables

To more directly test whether divergences in life course structure drove changes in attitude stability, we include a set of covariates capturing demographic transitions. These include 1) whether a respondent is currently enrolled in school; 2) whether a respondent has ever been married; and 3) whether a respondent have ever had a child. We also include an indicator for whether a respondent reported completing college.

@@

Table XXX presents a comparison of the Bayesian Information Criteria for models allowing for random effects for different coefficients. The best-fitting model, according to the BIC, allows random effects for duration, time, and age group, but does not allow them for any higher-order interaction. The best-fitting model also omits the third-order interaction between time, age group, and duration of observation. In other words, while there is evidence that age groups changed the amount they changed over time, and that the effect of duration changed over time, there is no evidence that groups experienced differnet rates of change in the rate of durable change.

@@

The intercept term in Model 1 suggests that, on average, survey respondents in the 18-25 age group are expected to vary by almost a fifth of the response scale on average (18 points) in the most recent years of the survey, controlling for duration between observations, on any particular question. This is a rather high level of variance, but it is consistent with previous studies that find that on any particular question survey respondents tend to make large wave-to-wave changes, or that survey questions about political attitudes have high levels of measurement error [@ansolabehere2009; @converse1965]. At the same time, random-effect variance around this main intercept term is quite large ($\sigma_{\beta_0} = 5.9220$). On several questions, respondents are expected to change their responses by about 40 percent of the scale between observations, while on others almost no wave-to-wave change is expected, meaning people tend to respond with the same answer over time.

Duration between observations is positively associated with wave-to-wave absolute change. For every additional year between responses, the expected absolute change for respondents increases 1.1 points. As suggested above, there is no evidence to suggest that respondents in different age groups vary systematically in this kind of change. Variance around this main effect is relatively small (s.d. = 0.36), meaning each one-year duration only increases the amount changed by between .5 and 1.5. Given the comparatively high level of wave-to-wave change independent of duration (18 points), our results broadly suggest that most wave-to-wave change is non-persisting, consistent with findings from other recent work [@kiley2020; @lersch2023].

@@

The intensification of early adulthood hypothesis (H3) generally suggests that the youngest respondents should make more change over time in more recent cohorts than their counterparts from earlier cohorts did when they were the same ages. The effect of year of observation in decades for people in the 18-25 age group ($\beta = 0.14345$, $t = 0.754$) is substantively small and not statistically significant. On average, recent cohorts in this age group make approximately the same amount of wave-to-wave change in recent years as earlier cohorts did, accounting for the kinds of questions they were asked and time between waves. However, question-level random effects vary considerably around this main effect, with a standard deviation of about 1.4316. This suggests that the overall flat effect obscures quite large question-level differences, with some questions becoming more stable and others becoming less stable. We return to consider these question-specific trends later.

Next, the extended early adulthood framework (H1) suggests that we should see a closing of the gap between people in the 18-25 and 26-33 age groups over time. The interaction between year of observation and the 26-33 age category is small and not statistically significant ($\beta = -0.03560$, $p=-0.361$), suggesting that people in this age group have also not become any more or less likely to make changes of opinion since the earliest panels, and have also not become any more or less similar to the youngest respondents, and that significant changes in the social environment have not made more recent cohorts any more likely to make attitude changes than earlier cohorts were when they were in this age range.

However, it is important to understand the full picture of change over time before drawing substantive conclusions. Table XXX shows that, in contrast to people in the 18-25 and 26-33 age range, respondents in older age groups do show a notable time trend -- people in this age range become more stable opinion holders over time. To more clearly illustrate overall trends, Figure XXX plots the predicted time level of change for respondents in each age category over time.

@@

In the earliest panels, the oldest respondents (65 and over) are substantially less stable opinion holders than younger respondents, while the remaining three age groups (18-25, 26-33, and 34-64) appear to change at approximately equal rates. Over time, middle aged respondents (34-64) and the oldest respondents (65+) become more stable. Because, as noted previously, respondents in the 18-25 and 26-33 age ranges demonstrate similar levels of opinion change over time, they become the least stable respondents by the end of the panel.

@@

There are no questions where members of one group become substantially more stable while another becomes less stable.

>>>>

Adolescence and young adulthood are frequently identified in social science literature as "impressionable years" -- a stage of the life course in which people are more susceptible to broad social currents, when their opinions are more malleable, and when they form the core ideas and attitudes that shape subsequent decision making [@krosnick1989; @alwin1991]. In support of this idea, social science research regularly finds that people in these life stages are more likely to change their political preferences, social attitudes, and cultural tastes than people in other life stages [@sears2013; @kiley2020]. This pattern, in which young people show greater openness to cultural change but become more resistant over time, underpins the dynamics of cohortization that are frequently invoked in larger models of social and cultural change [@ryder1965; @mannheim1952].

Why people in this life stage appear to show more frequent opinion change relative to older people, however, is unclear. Explanations in the social sciences tend to highlight the unique social features of these early life stages such as high rates of mobility between social contexts, participation in educational institutions, and cultural expectations about pursuing "new experiences," arguing that these conditions shape how people are exposed to new sources of information and their ability to update opinions in the face of new information [@alwin1991; @danigelis2007]. Other related explanations focus on the social-structural features of middle age that might stabilize attitudes [@eaton2009; @visser2004]. At the same time, work in biology, psychology, and cultural evolution finds that peoples' cognition changes in ways that might make them more resistant to new information as they age [CITATION], suggesting the differences between early adults and the rest of the population might be more rooted in biological and cognitive mechanisms than social mechanisms.

While both social and cognitive/biological mechanisms likely play a role in facilitating opinion change during early adulthood, the relative contributions of each are unclear, and clarifying their relative contributions is important for understanding the mutually reinforcing processes of social change and cultural evolution. If openness to cultural change in early adulthood is principally a function of the social structure of life during these ages, then delays in the timing and ubiquity of life course transitions like marriage and childbearing, and increases in college-going, could have significant effects on the broader process of macro-social change. Similarly, social changes such as the deinstitutionalization of marriage [@cherlin2004; @cherlin2021; @robbins2022] that decrease distinctions between life course stages might make older people more open to change in more recent eras than they were in earlier eras, diminishing the importance of "cohorts" for social change and increasing the importance of intrapersonal change over time [@kiley2020; @vaisey2016]. If the mechanisms are principally cognitive and biological, rather than social, then these broader changes are unlikely to affect rates of social change in the same way.

In this paper, we explore how rates of opinion change in different age groups have changed over the past 70 years in response to broader social structure and changes in the timing of life course events. Since the 1950s, the timing of life-course transitions such as completing schooling, entering the workforce, getting married, leaving one's parents' home, and having children have changed dramatically, with most of these events occurring at later ages for more recent cohorts [@buchmann1989; @bruckner2005]. The "early adulthood" stage of life has become more detached from family and other role obligations, giving people in this life stage more freedom to explore new ideas and lifestyle choices [@rosenfeld2009]. These broad social changes lengthen of and intensify this stage of the life course in ways that potentially have consequences for the rate at which people undergo cultural changes.

Drawing on data for 77 questions asked multiple times across 12 panel surveys dating from the 1950s through the 2020s, we model the amount of change people make over time in social and political attitudes as a function of the year they were observed and their age at the time. Consistent with the "impressionable years" hypothesis, we find that members of the youngest age groups (18-25 and 26-33) have more variable opinions than older respondents. However, these differences are not constant over time. They emerged over the second half of 20th Century as middle-aged and older respondents came to report more stable opinions over time, while early adult respondents' reported opinions continued to change at roughly the same rate over time.

Our findings complicate the narrative about social change and cultural change and raise questions about which social forces emerged to stabilize attitudes in middle age for more recent cohorts. Life-course stages can potentially explain the divergence of opinion stability between the youngest age groups and older age groups over time, but they cannot explain the overall trajectory of increased stability.

@@

These aggregate models of social imprinting have implications for the kinds of patterns we should observe in within-person opinion change over time. The "impressionable years" hypothesis -- that people demonstrate more opinion change early in life followed by a period of relative stability, especially in middle age -- has often been tested against other theories of opinion change over the life course. The most common alternatives are an "aging stability" or "increasing persistence" hypothesis that argues that openness to change gradually declines over time and a "life stages" hypothesis, in which people are open to attitude change both early in life and later in life, with stability principally happening in middle ages [@visser2004a]. While different attitudes occasionally follow different patterns, the impressionable years hypothesis has generally been shown to be true for broad political attitudes including political ideology, partisan identification, interest in politics, and specific policy preferences [@krosnick1989; @alwin1991; @sears1999; @prior2010]; social attitudes around race, gender, and sexuality [@danigelis2007]; as well as things like personality traits [@wagner2019] and cultural tastes [\@]. More generally, recent research has found that, with a few notable exceptions of high-salience events such as national shifts around gay rights, durable change of social opinion is quite rare among U.S. adults [@kiley2020], and Vaisey and Lizardo [-@vaisey2016] find that cohort, the year of a person's birth, explains much more variance than period, the year of observation, for most items in the General Social Survey, suggesting a strong impressionable years pattern.

@@

While we have focused on changes in the timing and structure of life course stages over the past 70 years as our primary focus, the past 70 years have also seen significant shifts in the broader social structure and cultural environment that might affect attitude stability independent of changes in life course structure. We briefly consider three broader changes here.

First, the number of people pursuing and completing higher education increased significantly over the window studied here [@schofer2005; @hout2012]. Existing work suggests that people with higher levels of education are more likely to express the same opinion over time in survey contexts [@alwin1991a; @alwin2007]. A population with higher levels of education, on average, might be more likely to hold stable opinions regardless of transformations in life course stages.

Second, technological changes have dramatically lowered barriers to accessing information, including information about social and political affairs that might facilitate stable opinion holding. Broadcast and cable television, followed by the internet and social media, have provided new channels for elites to communicate messages to the public and for people to acquire informaiton that makes opinion-holding easier.

Third, the past 70 years have seen extensive elite polarization around political and social affairs, as well broader public partisan alignment [@fiorina2008; @baldassarri2008]. People now receive clearer signals from cultural elites, including politicians, the news media, and others, about which beliefs and identities "go together" in the population, resulting in what DellaPosta (2020) calls an "oil spill" of polarization [@dellaposta2020]. This issue and identity alignment might make it easier for the public to identify what beliefs people like them "should" hold and hold them consistently [@zaller1992]. Geographic and social sorting into lifestyle "enclaves" also reinforce these processes, continually providing clearer signals about which opinions people should hold [@dellaposta2015].

Broadly speaking, these three forces suggest that opinions should broadly become more stable over time, regardless of life-course stages. Additionally, opinions on certain issues might become more solidified the longer issues remain in the public, as social movements and organizations become solified around them.

@@

Political and Social Attitudes: Our general question is whether, when presented with the same question during the same age range, more recent cohorts will demonstrate more change over time than members of earlier cohorts. Analyzing this question requires data on opinion and attitude reports over time for members of different cohorts at different time points, with respondents coming from the same general population. To construct such a data source, we combine several panel surveys conducted by the American National Election Studies and the General Social Survey spanning about 60 years. Both the ANES and GSS are designed to generalize to the non-institutionalized U.S. adult (18+) population, and therefore can be thought of as tapping the same general population. We included a total of 12 panels in our search: ANES panels from 1956-60, 1972-76, 1980, 1990-92, 1992-97, 2000-04, 2016-20, and 2020-22; and GSS panels from 2006-10, 2008-12, 2010-14, and 2016-20. We evaluated an additional panel conducted by the ANES in 2008-10, but no questions overlapped this panel and any other panel.

For reasons explained below, we identified questions that were asked in at least three different panels with at least one of those panels asking some respondents to answer the question at three or more separate time points. Because three of the GSS panels overlap considerably (2006-10, 2008-12, and 2010-14) and contain the same set of questions, we did not treat these as separate panels for the purposes of generating three distinct time points. In other words, questions needed to appear in these panels and an ANES panel to be included in our analysis. When question wording changed across panels, we harmonized response options to the question structure with the fewest response options.

Our search produced a total of 77 questions tapping a broad range of topics including political and ideological identification, general government policy preferences, preferences for federal spending on different priorities, views on government efficacy, views on social issues and social change, and sentiment toward various political and social groups. For reasons elaborated below, we scaled all items to have a minimum of 0 and a maximum of 100.

@@

### Demographic Measures

The central question of our analysis pertains to age, and as such we classify respondents into a set of age brackets depending on their age when they answered survey questions. Given our theoretical questions, we classify people into four age categories: 18-25, 26-33, 34-64, and 64-80. Because of differences in both question structure and age structure across panels, we drop respondents who are over 80.

We are also interested in the effect of having undergone a set of life-course transitions on attitude stability to more directly test whether divergences in life course structure drove changes in attitude stability. To this end, we include a set of indicator variables capturing whether people have undergone certain demographic transitions. These include 1) whether a respondent is currently enrolled in school; 2) whether a respondent has ever been married; and 3) whether a respondent has ever had a child. We also include an indicator for whether a respondent reported completing college to separately explore whether higher rates of college completion explain changes over time across age groups.

@@

When questions are observed in multiple panels, differences in the amount of change we observe could be attributable to either changes in the population's capacity to report stable attitudes over time (our main theoretical interest) or to the structural features of the survey administration. Specifically, variation in the amount of change people makes in their responses observed over time might be due to variation in the duration between waves, with longer duration between observations leading to more wave-to-wave change as people experience more shocks that should shift their opinions. In our data, panel duration varied from less than one year (the 1980 ANES Major Panel Study) to about five years (the 1992-97 ANES), and duration between questions ranged from a few days (for questions asked in both pre- and post-election surveys in the ANES panels) to about five years (questions asked in 1992 and 1997), with peaks around 2 years and 4 years.

@@

In this equation, $d_{i}$ captures the duration of time between $t_1$ and $t_2$ in years for each pair of observations by the same respondent. Importantly, respondents might have multiple observation pairs for each panel, depending on the number of times a question was asked. In the 1972-76 ANES panel, respondents were asked their partisan identification on three occasions: in the 1972 pre-election survey, in the 1974 pre-election survey, and the 1976 pre-election survey, each about two years apart. If a person participates in all three waves, they have three observations of $|y_{2i}-y_{1i} |$ and $d_i$: 1972 and 1974 (about 2 years), 1974 and 1976 (about 2 years), and 1972 and 1976 (about 4 years).

Variation in duration between observations within panels lets us separate two kinds of change: 1) change attributable to durable shifts of opinions over time that can be associated with duration between observations and 2) change that is principally random, or unrelated to the duration between observations. If respondents have a stable underlying partisan identification they report with random error, and no real change occurs, then in expectation, duration should be unrelated to absolute change in responses over time, and $\beta_1 = 0$. In this formulation, $\beta_0$ captures the expected absolute change in responses over time when the duration between waves is 0, or, more intuitively, the random deviation between waves that is unrelated to durable change. Conversely, $\beta_1$ captures the expected change in absolute difference between waves for a one-unit (year) change in duration of observation, or the amount of absolute change we would expect over-and-above the change driven by random noise ($\beta_1$). For example, if a respondent shifts from "strong Democrat" (point 0 on the scale) to "Independent" (point 3 on the scale) between 1972 and 1974, and stays at "Independent" for the 1976 wave, they will have differences of 3 and 0 for the two-year duration (an average of 1.5) and 3 for the four-year duration, generating a positive coefficient for $\beta_1$ if this kind of switching behavior is common in the populations.

Given our focus on age-based differences, we can extend equation to incorporate age groups as indicator variables:

$$ |y_{2i}-y_{1i} | = \beta_0 + \beta_1*d_i + \beta_2(a_{26-33}) + \beta_3(a_{34-64}) + \beta_4(a_{65+}) + \beta_5(d*a_{26-33}) + \beta_6(d*a_{34-64}) + \beta_7(d*a_{65+}) + \epsilon_i $$

@@

$\beta_0$ through $\beta_7$ retain their general meaning as outlined above, but apply to $t = 0$, which we set to be the most recent observation date (2016). $\beta_8$ estimates the change in the baseline rate of change for the reference group. $\beta_9$ captures change in the effect of duration of observation over time for people in the reference group (ages 18-25). Coefficients $\beta_{10}$ through $\beta_{12}$ and $beta_{13}$ capture how differences in the baseline rate of change and duration effects between the reference group and older age categories change over time.

@@

$$ \beta_{kj} = \gamma_k + \upsilon_{kj} $$ This model allows for each survey question to have its own set of coefficient estimates weighted by the amount of information we have about the particular question. Questions asked in many panels will contribute more to the overall sense of the aggregate trend in opinion stability.

In this framework, the first-level equation models the absolute difference in a pair of responses to the same question given by the same respondent, similar to what was presented above. These equations are nested at the level of the survey item. In practice, the model presented in Eq \### is likely too complicated in the face of the data limitations we face. As part of our analysis, we compare the fully elaborated model to more Criterion models using model-selection criteria, specifically the Bayesian Information Criterion. For example, it could be the case that little predictive information is gained by allowing questions to have separate slopes for the relationship between age categories and duration over time.

### Demographic Transitions

## Weights

@@

@@

Table XXX presents a comparison of the Akaike Information Criteria (AIC) and Bayesian Information Criteria (BIC) for the full model elaborated above (M1) and a series of more simplified model. The best-fitting model, according to the BIC, allows random effects for duration, time, and age group, but does not allow them for any of the higher-order interactions. The best-fitting model (M5) omits the third-order interaction between time, age group, and duration of observation and an the interaction between age group and duration. Further attempts to simplify the model, including dropping random effects for age categories and duration, resulted in increases in the BIC.

| Variables                      | M1       | M2       | M3       | M4       | M5       |
|------------|------------|------------|------------|------------|------------|
| Intercept                      | Y        | Y        | Y        | Y        | Y        |
| RE                             | Y        | Y        | Y        | Y        | Y        |
| Time                           | Y        | Y        | Y        | Y        | Y        |
| RE                             | Y        | Y        | Y        | Y        | Y        |
| Age Groups                     | Y        | Y        | Y        | Y        | Y        |
| RE                             | Y        | Y        | Y        | Y        | Y        |
| Duration                       | Y        | Y        | Y        | Y        | Y        |
| RE                             | Y        | Y        | Y        | Y        | Y        |
| Time \* Age Group              | Y        | Y        | Y        | Y        | Y        |
| RE                             | Y        | Y        | Y        | N        | N        |
| Time \* Duration               | Y        | Y        | Y        | Y        | Y        |
| RE                             | Y        | Y        | N        | N        | N        |
| Age Groups \* Duration         | Y        | Y        | Y        | Y        | N        |
| RE                             | Y        | Y        | N        | N        | N        |
| Age Groups \* Duration \* Time | Y        | N        | N        | N        | N        |
| RE                             | Y        | N        | N        | N        | N        |
|                                |          |          |          |          |          |
| BIC                            | 20714.39 | 20250.97 | 19996.61 | 19862.23 | 19845.22 |
| AIC                            | 19757.88 | 19675.82 | 19627.76 | 19643.42 | 19645.17 |

: Model fit comparison

@@

| Col1             | Col2  | Col3 | Col4 |
|------------------|-------|------|------|
| Fixed Effects    |       |      |      |
| Intercept        | 17.45 |      |      |
| Time (decade)    | -0.26 |      |      |
| Age(26-33)       | -1.60 |      |      |
| Age(34-64)       | -3.17 |      |      |
| Age(65+)         | -4.10 |      |      |
| Duration (year)  | 1.13  |      |      |
| Time\*Age(26-33) | -0.09 |      |      |
| Time\*Age(34-64) | -0.54 |      |      |
| Time\*Age(65+)   | -1.33 |      |      |
|                  |       |      |      |
| Random Effects   | S.D.  |      |      |
| Intercept        | 5.35  |      |      |
| Time             | 1.14  |      |      |
| Duration         | 0.37  |      |      |
| Age(26-33)       | 0.89  |      |      |
| Age(34-64)       | 1.81  |      |      |
| Age(64+)         | 2.92  |      |      |

: Coefficient Estimates

The intercept term in Model 1 suggests that, on average, survey respondents in the 18-25 age group are expected to vary by almost a fifth of the response scale on average (18 points) in the most recent years of the survey, controlling for duration between observations, on any particular question. This is a rather high level of variance, but it is consistent with previous studies that find that on any particular question survey respondents tend to make large wave-to-wave changes, or that survey questions about political attitudes have high levels of measurement error [@ansolabehere2009; @converse1965]. At the same time, random-effect variance around this main intercept term is quite large ($\sigma_{\beta_0} = 5.9220$). On several questions, respondents are expected to change their responses by about 40 percent of the scale between observations, while on others almost no wave-to-wave change is expected, meaning people tend to respond with the same answer over time with low levels of measurement error or random change.

Duration between observations is positively associated with wave-to-wave absolute change. For every additional year between responses, the expected absolute change for respondents increases 1.1 points. As suggested above, there is no evidence to suggest that respondents in different age groups vary systematically in this kind of change. Random effect variance around this main effect is relatively small (s.d. = 0.36), meaning each one-year duration only increases the amount changed by between .5 and 1.5. Given the comparatively high level of wave-to-wave change independent of duration (18 points), our results broadly suggest that most wave-to-wave change is non-persisting, consistent with findings from other recent work [@kiley2020; @lersch2023].

@@

Coefficients in Model 1 allow us to test several of the hypotheses outlined above. The intensification of early adulthood hypothesis (H3) generally suggests that the youngest respondents should make more change over time in more recent cohorts than their counterparts from earlier cohorts did when they were the same ages. The effect of year of observation in decades for people in the 18-25 age group ($\beta = 0.14345$, $t = 0.754$) is substantively small and not statistically significantly different from 0. On average, recent cohorts in this age group make approximately the same amount of wave-to-wave change in recent years as earlier cohorts did, accounting for the kinds of questions they were asked and time between responses. However, question-level random effects vary considerably around this main effect, with a standard deviation of about 1.43. This suggests that the overall flat trend obscures quite large question-level differences, with some questions becoming more stable and others becoming less stable. We return to consider these question-specific trends later.

Next, the extended early adulthood framework (H1) suggests that we should see a closing of the gap between people in the 18-25 and 26-33 age groups over time. The interaction between year of observation and the 26-33 age category is small and not statistically significant ($\beta = -0.03560$, $p=-0.361$), suggesting that people in this age group have not become any more or less likely to make changes of opinion since the earliest panels, have not become any more or less similar to the youngest respondents.

However, it is important to understand the full picture of change over time before drawing substantive conclusions. Table XXX shows that, in contrast to people in the 18-25 and 26-33 age range, respondents in older age groups (34-64 and 65+) do show a notable time trend -- people in these age ranges become more stable opinion holders over time. To more clearly illustrate overall trends, Figure XXX plots the predicted time level of change for respondents in each age category over time.

@@

In the earliest panels, the oldest respondents (65 and over) are substantially less stable opinion holders than younger respondents, changing by about 5 points more than members of the other groups, while the remaining three age groups (18-25, 26-33, and 34-64) appear to change at approximately equal rates. Over time, middle aged respondents (34-64) and the oldest respondents (65+) become more stable, with the latter becoming almost 10 points more stable on average. Because, as noted previously, respondents in the 18-25 and 26-33 age ranges demonstrate similar levels of opinion change over time, they become the least stable respondents by the end of the panel.

## Life Course Shifts

While coefficient estimates for the early adult age groups do not suggest that they have become more variable opinion holders over time, the increase in stability for middle-aged and older respondents suggests that broad social forces might counteract life-course the effect of transitions.

To evaluate the specific effect of life-course changes, we attempt to control for these transitions. We re-estimate Model \### above, this time incorporating a series of indicator variables for whether people were students, had children, and were married. We also include an indicator variable for whether the respondent completed college to capture the effect of ... .

Using these estimates, we calculate the expected value of change had life-course transitions occurred at roughly the same rates.

@@

As noted above, the overall trajectory of increasing attitude stability masks significant heterogeneity across questions. To more clearly illustrate these differences in trends, Figure XXX plots the estimated trajectories for each item for respondents in the youngest (18-25) and middle-aged (34-64) age categories. Items are ordered by their predicted time trends.

```{r}
slopes <- as.data.frame(ranef(m4.5)$name) %>%
  rownames_to_column(var = "name") %>%
  mutate(slope = dec_diff) %>%
  select(name, slope)
